Supp mat: time series paper

### 2017-08-16 writing/manuscript_draft.Rmd
small changes in refs of figures

@@

@@

We examined a database of `r nrow(pop_info)` separate population time series representing `r length(table(pop_info$Binomial))` species. This database consists of species with a variety of life history characteristics. We limited our analyses to populations with at least 35 years of continuous sampling.

@@

## Questions in the supplmentary material

In the supplementary material, we explore several additional questions related to statistical power and estimating $T_{min}$. First, we showed that other metrics besides estimating the slope coefficient from linear regression could be used. More specifically, we show how the same approach described here can be used to determine the minimum time required to estimate a population's geometric growth rate (Figs. \ref{fig:growth_rate},\ref{fig:min_time_growth_dist}). We also explored how the simulation approach can be applied to more biologically-realistic population models (Fig. \ref{fig:shark_example}). More specifically, we determine the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014]. We found that over 27 years of continuous monitoring were needed in this particular scenario (Fig. \ref{fig:shark_example}. Lastly, we determine the minimum time required to estimate long-term trends according to generalized additive models, instead of the simple linear models used here (Fig. \ref{fig:gam_example}). We found that although we obtain a similar distribution of minimum times required for GAMs, the minimum time required for GAMs is on average shorter than for linear regression (Fig. \ref{fig:min_time_dist_gam}). We expand on these results in the supplementary material.

@@

Here, we focus on an empirical approach to estimate the minimum time series length required to assess changes in abundance over time. We examined long time series and then subsampled each to determine the minimum time required to achieve a desired level of significance and power for linear regression. Statistical power is important as it provides on information as to the necessary samples required to determine a significant trend [@Legg2006]. We found that for `r nrow(pop_info)` populations, at least 10-15 years of continuous monitoring were necessary (Fig. \ref{fig:min_time_dist}). This time frame is in line with past work on a short-lived snail species [@Rueda-Cediel2015] and a long-lived whale species [@Gerber1999]. In line with theoretical predictions [@Rhodes2011], we also found $T_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation (Fig. \ref{fig:correlates}). Contrary to our prior expectations, we also found that $T_{min}$ did not correlate with any biological variables of interest (Fig. \ref{fig:biological_correlates}). We initially hypothesized that species with longer lifespans or generation times may require a longer sampling period. Our result could have been a result of at least two effects. First, the database we used may not include a diverse enough set of species with different life history traits. Second, the question we poised, whether a population is increasing or decreasing, was specifically concerned with trends in population density over time. Therefore, life-history characteristics may be more important for other questions (e.g. extinction risk) where estimates minimum time required are needed.

@@

ERW was partially supported by a National Science Foundation Graduate Fellowship. We would like to thank members of the Ecological Theory group at the University of California, Davis for their insight. We would also like to thank T. Dallas and E. Malcolm-White<!--X anonymous reviewers -->for their helpful comments.

>>>>

We also explored how the simulation approach can be applied to more biologically-realistic population models (Fig. \ref{fig:shark_example}). More specifically, we determine the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014]. We found that over 27 years of continuous monitoring were needed in this particular scenario (Fig. \ref{fig:shark_example}). Similar to the simulation approach decribed above, the minimum time required for the lemon shark population was strongly dependent on model parameters.

@@

In the supplementary material, we show how the same approach described here can be used to determine the minimum time required to estimate a population's geometric growth rate (Figs. \ref{fig:growth_rate},\ref{fig:min_time_growth_dist}). We also determine the minimum time required to estimate long-term trends according to generalized additive models, instead of the simple linear models used here (Fig. \ref{fig:gam_example}). We found that although we obtain a similar distribution of minimum times required for GAMs, the minimum time required for GAMs is on average shorter than for linear regression (Fig. \ref{fig:min_time_dist_gam}).

@@

We examined a database of `r nrow(pop_info)` separate population time series representing `r length(table(pop_info$Binomial))` species. This database consists of species with a variety of life history characteristics. We limited our analyses to populations with at least 35 years of continuous sampling. We examined the minimum time required to estimate long-term trends via linear regression.

@@

<!--Integrate with other sections
## Questions in the supplmentary material

In the supplementary material, we explore several additional questions related to statistical power and estimating $T_{min}$. First, we showed that other metrics besides estimating the slope coefficient from linear regression could be used. More specifically, we show how the same approach described here can be used to determine the minimum time required to estimate a population's geometric growth rate (Figs. \ref{fig:growth_rate},\ref{fig:min_time_growth_dist}). We also explored how the simulation approach can be applied to more biologically-realistic population models (Fig. \ref{fig:shark_example}). More specifically, we determine the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014]. We found that over 27 years of continuous monitoring were needed in this particular scenario (Fig. \ref{fig:shark_example}. Lastly, we determine the minimum time required to estimate long-term trends according to generalized additive models, instead of the simple linear models used here (Fig. \ref{fig:gam_example}). We found that although we obtain a similar distribution of minimum times required for GAMs, the minimum time required for GAMs is on average shorter than for linear regression (Fig. \ref{fig:min_time_dist_gam}). We expand on these results in the supplementary material.
-->

@@

Here, we focus on an empirical approach to estimate the minimum time series length required to assess changes in abundance over time. We examined long time series and then subsampled each to determine the minimum time required to achieve a desired level of significance and power for linear regression. Statistical power is important as it provides on information as to the necessary samples required to determine a significant trend [@Legg2006]. We found that for `r nrow(pop_info)` populations, at least 10-15 years of continuous monitoring were necessary (Fig. \ref{fig:min_time_dist}). This time frame is in line with past work on a short-lived snail species [@Rueda-Cediel2015] and a long-lived whale species [@Gerber1999]. In line with theoretical predictions [@Rhodes2011], we also found $T_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation (Fig. \ref{fig:correlates}). Contrary to our prior expectations, we also found that $T_{min}$ did not correlate with any biological variables of interest (Fig. \ref{fig:biological_correlates}). We initially hypothesized that species with longer lifespans or generation times may require a longer sampling period. Our result could have been a result of at least two effects. First, the database we used may not include a diverse enough set of species with different life history traits. Second, the question we poised, whether a population is increasing or decreasing, was specifically concerned with trends in population density over time. Therefore, life-history characteristics may be more important for other questions (e.g. species extinction risk) where estimates minimum time required are needed.

@@

ERW was partially supported by a National Science Foundation Graduate Fellowship. We would like to thank members of the Ecological Theory group at the University of California, Davis for their insight. We would also like to thank T. Dallas and E. Malcolm-White <!--X anonymous reviewers -->for their helpful comments.

### 2017-08-16 writing/manuscript_draft.tex
small changes in refs of figures

@@

We examined a database of 878 separate population time series
representing 478 species. This database consists of species with a
variety of life history characteristics. We limited our analyses to
populations with at least 35 years of continuous sampling.

@@

\subsection{Questions in the supplmentary
material}\label{questions-in-the-supplmentary-material}

In the supplementary material, we explore several additional questions
related to statistical power and estimating \(T_{min}\). First, we
showed that other metrics besides estimating the slope coefficient from
linear regression could be used. More specifically, we show how the same
approach described here can be used to determine the minimum time
required to estimate a population's geometric growth rate (Figs.
\ref{fig:growth_rate},\ref{fig:min_time_growth_dist}). We also explored
how the simulation approach can be applied to more
biologically-realistic population models (Fig. \ref{fig:shark_example}).
More specifically, we determine the minimum time required to estimate
long-term population trends using a stochastic, age-structured model of
lemon shark population dynamics in the Bahamas (White, Nagy, and Gruber
2014). We found that over 27 years of continuous monitoring were needed
in this particular scenario (Fig. \ref{fig:shark_example}. Lastly, we
determine the minimum time required to estimate long-term trends
according to generalized additive models, instead of the simple linear
models used here (Fig. \ref{fig:gam_example}). We found that although we
obtain a similar distribution of minimum times required for GAMs, the
minimum time required for GAMs is on average shorter than for linear
regression (Fig. \ref{fig:min_time_dist_gam}). We expand on these
results in the supplementary material.

@@

Here, we focus on an empirical approach to estimate the minimum time
series length required to assess changes in abundance over time. We
examined long time series and then subsampled each to determine the
minimum time required to achieve a desired level of significance and
power for linear regression. Statistical power is important as it
provides on information as to the necessary samples required to
determine a significant trend (Legg and Nagy 2006). We found that for
878 populations, at least 10-15 years of continuous monitoring were
necessary (Fig. \ref{fig:min_time_dist}). This time frame is in line
with past work on a short-lived snail species (Rueda-Cediel et al. 2015)
and a long-lived whale species (Gerber, DeMaster, and Kareiva 1999). In
line with theoretical predictions (Rhodes and Jonzen 2011), we also
found \(T_{min}\) was strongly correlated with the trend strength,
variability in population size, and temporal autocorrelation (Fig.
\ref{fig:correlates}). Contrary to our prior expectations, we also found
that \(T_{min}\) did not correlate with any biological variables of
interest (Fig. \ref{fig:biological_correlates}). We initially
hypothesized that species with longer lifespans or generation times may
require a longer sampling period. Our result could have been a result of
at least two effects. First, the database we used may not include a
diverse enough set of species with different life history traits.
Second, the question we poised, whether a population is increasing or
decreasing, was specifically concerned with trends in population density
over time. Therefore, life-history characteristics may be more important
for other questions (e.g.~extinction risk) where estimates minimum time
required are needed.

>>>>

We also explored how the simulation approach can be applied to more
biologically-realistic population models (Fig. \ref{fig:shark_example}).
More specifically, we determine the minimum time required to estimate
long-term population trends using a stochastic, age-structured model of
lemon shark population dynamics in the Bahamas (White, Nagy, and Gruber
2014). We found that over 27 years of continuous monitoring were needed
in this particular scenario (Fig. \ref{fig:shark_example}). Similar to
the simulation approach decribed above, the minimum time required for
the lemon shark population was strongly dependent on model parameters.

@@

We examined a database of 878 separate population time series
representing 478 species. This database consists of species with a
variety of life history characteristics. We limited our analyses to
populations with at least 35 years of continuous sampling. We examined
the minimum time required to estimate long-term trends via linear
regression.

@@

@@

Here, we focus on an empirical approach to estimate the minimum time
series length required to assess changes in abundance over time. We
examined long time series and then subsampled each to determine the
minimum time required to achieve a desired level of significance and
power for linear regression. Statistical power is important as it
provides on information as to the necessary samples required to
determine a significant trend (Legg and Nagy 2006). We found that for
878 populations, at least 10-15 years of continuous monitoring were
necessary (Fig. \ref{fig:min_time_dist}). This time frame is in line
with past work on a short-lived snail species (Rueda-Cediel et al. 2015)
and a long-lived whale species (Gerber, DeMaster, and Kareiva 1999). In
line with theoretical predictions (Rhodes and Jonzen 2011), we also
found \(T_{min}\) was strongly correlated with the trend strength,
variability in population size, and temporal autocorrelation (Fig.
\ref{fig:correlates}). Contrary to our prior expectations, we also found
that \(T_{min}\) did not correlate with any biological variables of
interest (Fig. \ref{fig:biological_correlates}). We initially
hypothesized that species with longer lifespans or generation times may
require a longer sampling period. Our result could have been a result of
at least two effects. First, the database we used may not include a
diverse enough set of species with different life history traits.
Second, the question we poised, whether a population is increasing or
decreasing, was specifically concerned with trends in population density
over time. Therefore, life-history characteristics may be more important
for other questions (e.g.~species extinction risk) where estimates
minimum time required are needed.
